Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-07-29 paper/main.tex
move section to supplement
write about individual treatment of cells

%###################################################################################################
\section{Application to \textit{E.Coli.} growing on Agar (TODO)}
% --------------------------------------------------------------------------------------------------
\subsection{Data Preparation}
TODO: Decide if this should be Supplement
\begin{enumerate}
    \item Have masks of data; reference image segmantation tools; or do it by hand
    \item ensure that colors are matching for bacteria between time-steps
    \item clean borders of images if needed
    \item know domain size; otherwise use pixels
\end{enumerate}

% --------------------------------------------------------------------------------------------------
\subsection{Non-Individual Treatment}
% --------------------------------------------------------------------------------------------------
\subsection{Individual Treatment}
\label{subsec:parameter-estimation-individual-treatment}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-individual.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-average.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/parameter-distribution.pdf}%
    \caption{}
    \label{fig:estimated-growth-rates}
\end{figure}
% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
We will now focus on using the presented methods for parameter estimation via microscopic images.
Our initial state is given by TODO~\cite{}.% TODO
We use our extraction algorithm to obtain the initial positions of our cells.
From there, we supply the model with parameters such that we can make a prediction for the final
positions of the bacteria.
We chose a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our cost
function.
Another important detail is the individuality of the agents.
In our case, this means that each agent might have different parameters associated to it.
For this dataset, we assumed that all interaction parameters which specify the interaction
potential are identical across all agents but growth rates and the physical radii of the cells are
individual and we also optimize them.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube sweep across the parameter space and
performed a local minimization with the Nelder-Mead method (TODO citations).% CITATIONS

@@

>>>>

%###################################################################################################
\section{Application to \textit{E.Coli.} growing on Agar (TODO)}

With the previous works of extracting the positions (see section~\ref{section:data-extraction}) for
individual cell-agents and the methods to estiamte parameters of our agents (see section~\ref{section:parameter-estimation}), we are now ready to apply them to showcase their usefulness.

TODO explain what data we are using

% --------------------------------------------------------------------------------------------------
\subsection{Individual Treatment}
We have already suspected (see table~\ref{table:simulation-aspects}) that in order to properly
estimate parameters of our system, we need to treat every cell-agent individually, thus assigning
parameters individually to them.
This means that we increase the total number of parameters that needs to be sampled within an
optimization routine.
Given $N_c$ cells and $N_p$ parameters per cell and $N_u$ unique parameters (where $N_u\leq N_p$),
the total amount of parameters is

\begin{equation}
    N_{p,\text{total}} = N_u + N_c (N_p - N_u) = N_C N_p - (N_c - 1) N_u.
\end{equation}

It is clear, that we should avoid to needlessly assign every parameter individually since the space
which is needed to be sampled scales exponentially with the dimension ie. the number of parameters.
However, we will now present data that indicate that an individual treatment is necessary.
Figure~\ref{fig:estimated-growth-rates} shows the rod-lengths of individual bacteria which have been
calculated from the extracted positions.
We fit a delayed exponential function to them which is defined by

\begin{equation}
    x(t) =
    \left\{\begin{array}{ll}
            x_0 \exp(\lambda (t-t_0)) & t\geq t_0\\
            x_0 & \text{ else}
    \end{array}\right.
\end{equation}

and where $\lambda$ is the growth rate $t_0$ the delay.
The fits to invidiual growth-curves of figure~\ref{fig:estimated-growth-rates} (A) result in good
matches and so does the fit of subfigure (B).
However, the latter exhibits large amounts of uncertainty within the data which is also illustrated
in subfigure (C).
There, we can see much more clearly the difference between considering invidiually assigned
parameters and global ones.
It is clear from this observation that in order for us to accurately estimate the parameters of this
system, we need to treat the bacteria individually.

The individual treatment of the bacteria allows us in principle to estimate the distribution of the
parameters directly without having to resort to any assumptions on its form.
This is true for any parameter, not just the growth rates, since we aim to estimate all mechanical
parameters of our cellular model.

\textbf{TODO is this discussion?}\\
The total parameter space $P$ is a product of all single spaces $P_i$ in which the parameters
$p_i$ live $P=P_0\times P_1\times\dots$ but some of these parameters share the same space (let's
assume $p_0,p_1\in \tilde{P}$).
This leaves the open question as to weather the optimization algorithm could be modified to leverage
this knowledge and obtain increased performance.
For instance, it might be valuable to initially consider individual parameters as globally defined
and start treating the individually after a certain amount of optimization steps.

% \begin{enumerate}
%     \item motivate individual treatment of cells
%     \item pro: get at distributions of parameters; even more than growth rates
%     \item pro: probably good idea to do this in this case, since more or less required (see figure)
%     \item con: computationally more demanding
%     \item can we improve the optimization algorithm to somehow "know" that some parameters are for
%         different cells and thus do not "interact" as other parameters with each other?
%         ie. they are part of the same subspace.
% \end{enumerate}

\label{subsec:parameter-estimation-individual-treatment}
\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-individual.pdf}%
        \node at (0, 1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-average.pdf}%
        \node at (0, 1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/parameter-distribution.pdf}%
        \node at (0, 1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}%
    \caption{
        % TODO use same y-axis across the plots
        (A) The subfigure shows the time-evolution of the rod-lengths for every individual
        bacterium.
        These growth curves are fitted individually with delayed exponential function.
        (B) The datapoints are created by averaging the rod-lengths shown in subfigure (A) and then
        fitted with an exponential function.
        The uncertainty band is the standard deviation of the averaged datapoints.
        (C) Blue dots show the combination of estimated growth rates and delays for individual
        bacteria while the orange dot corresponds to the average.
        Ellipses around the datapoints are taken from the uncertainty of the fits.
        Time is given in units of frames of the movie from which the data have been extracted.
    }
    \label{fig:estimated-growth-rates}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
We will now focus on using the presented methods for parameter estimation via microscopic images.
Our initial state is given by TODO~\cite{}.% TODO
We use our extraction algorithm to obtain the initial positions of our cells.
From there, we supply the model with parameters such that we can make a prediction for the final
positions of the bacteria.
We chose a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our cost
function.
Another important detail is the individuality of the agents.
In our case, this means that each agent might have different parameters associated to it.
For this dataset, we assumed that all interaction parameters which specify the interaction
potential are identical across all agents but growth rates and the physical radii of the cells are
individual and we also optimize them.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube sweep across the parameter space and
performed a local minimization with the Nelder-Mead method (TODO citations).% CITATIONS

@@

%###################################################################################################
\section{Data Preparation}
\begin{enumerate}
    \item Have masks of data; reference image segmantation tools; or do it by hand
    \item ensure that colors are matching for bacteria between time-steps
    \item clean borders of images if needed
    \item know domain size; otherwise use pixels
\end{enumerate}
